Testing gene methylation enrichment analysis approaches using simulated data

### 2023-10-15 figs/GSE158422_simulate.Rmd
gsa sim

seed=100

# select gene sets to alter
gset_mod <- gset_selection(genesetdatabase=gs_symbols,seed=seed,gene_catalog=gene_catalog)

# select normal samples only
normal <- mval[,(1:(ncol(mval)/2)*2)]

# divide sets between hyper and hypomethylated
nset <- floor(length(gset_mod)/2)
set.seed(seed) ; gtup <-sample(gset_mod,48)
set.seed(seed) ; gtdn <- sample(setdiff(gset_mod,gtup),nset)
gup <- unname(unlist(gtup))
gdn <- unname(unlist(gtdn))

# divide samples between ctrl and case
set.seed(seed) ; ctrl <- sample(1:ncol(normal),18)
set.seed(seed) ; case <- sample(setdiff(1:37,ctrl),18)
mval_ctrl <- mval[,ctrl]
mval_case <- mval[,case]

# select probes

>>>>

seed=100
frac_genes=0.5
frac_probes=0.5
nsamples=10
normal_mval <- mval[,(1:(ncol(mval)/2)*2)]

incorp_dm <- function(genesetdatabase,myann,gene_catalog,mval,seed,frac_genes,frac_probes,groupsize) {
  # select gene sets to alter
  gset_mod <- gset_selection(genesetdatabase=genesetdatabase,seed=seed,gene_catalog=gene_catalog)
  # divide gene sets between hyper and hypomethylated
  nset <- floor(length(gset_mod)/2)
  set.seed(seed) ; gtup <-sample(gset_mod,nset)
  set.seed(seed) ; gtdn <- sample(setdiff(gset_mod,gtup),nset)
  gup <- unname(unlist(gtup))
  gdn <- unname(unlist(gtdn))
  # make probe-gene vector
  probe2gene <- strsplit(myann$UCSC_RefGene_Name,";")
  names(probe2gene) <- rownames(myann)
  probe2gene <- unlist(probe2gene)
  # select probes hypermethylated
  set.seed(seed) ; gup2 <- sample(gup,floor(length(gup)*frac_genes))
  pup <- names(probe2gene[which(probe2gene %in% gup2)])
  set.seed(seed) ; pup2 <- sample(pup,floor(length(pup)*frac_probes))
  # select probes hypomethylated
  set.seed(seed) ; gdn2 <- sample(gdn,floor(length(gdn)*frac_genes))
  pdn <- names(probe2gene[which(probe2gene %in% gdn2)])
  set.seed(seed) ; pdn2 <- sample(pdn,floor(length(pup)*frac_probes))
  # divide samples between ctrl and case
  ncols <- ncol(mval)
  maxgroupsize=floor(ncols/2)
  if ( groupsize > maxgroupsize ) { stop("groupsize cannot be larger than half the ncols of mval") }
  set.seed(seed) ; ctrl <- sample(1:ncols,groupsize)
  set.seed(seed) ; case <- sample(setdiff(1:ncols,ctrl),groupsize)
  mval_ctrl <- mval[,ctrl]
  mval_case <- mval[,case]
  # incorporate altered signals - change by +1 or -1
  mval_case[rownames(mval_case) %in% pup2,] <-  mval_case[rownames(mval_case) %in% pup2,] +1
  mval_case[rownames(mval_case) %in% pdn2,] <-  mval_case[rownames(mval_case) %in% pdn2,] -1
  mval2 <- cbind(mval_ctrl,mval_case)
  result <- list("mval"=mval2,"probes up"=pup2,"probes down"=pdn2,
    "genes up"=gup2,"genes down"=gdn2,
    "genesets up"=gtup,"genesets down"=gtdn)
  return(result)
}

## GSAMETH function

```{r,gsameth_func}

# limma
runlimma <- function(mval,design,myann) {
  fit.reduced <- lmFit(mval,design)
  fit.reduced <- eBayes(fit.reduced)
  dm <- topTable(fit.reduced,coef=ncol(design), number = Inf)
  dm <- merge(myann,dm,by=0)
  dm <- dm[order(dm$P.Value),]
  rownames(dm) <- dm$Row.names
  dm$Row.names=NULL
  return(dm)
}

This is how to use the function

```{r,rungsameth}

simgsa <- function(genesetdatabase,myann,gene_catalog,
  mval,seed,frac_genes,frac_probes,groupsize) {

  sim <- incorp_dm(genesetdatabase=gs_symbols,myann=myann,gene_catalog=gene_catalog,
    mval=mval,seed=seed,frac_genes=0.5,frac_probes=0.5,groupsize=groupsize)
  mval2 <- sim$mval
  ncols <- ncol(mval2)
  groupsize <- ncols/2
  ss <- data.frame(colnames(mval2))
  colnames(ss) <- "sample"
  ss$case <- c(rep(0,groupsize),rep(1,groupsize))
  d <- model.matrix(~ ss$case )
  dm3 <- runlimma(mval=mval2,design=d,myann=myann)
  pup3 <- rownames(subset(dm3,adj.P.Val<0.05 & logFC>0))
  pdn3 <- rownames(subset(dm3,adj.P.Val<0.05 & logFC<0))
  gsaup3 <- gsameth(sig.cpg=pup3, all.cpg=rownames(dm3), collection=gs_entrez)
  gsadn3 <- gsameth(sig.cpg=pdn3, all.cpg=rownames(dm3), collection=gs_entrez)
  gsig_up3 <- rownames(subset(gsaup3,FDR<0.05))
  gsig_dn3 <- rownames(subset(gsadn3,FDR<0.05))
  gtup <- names(sim[[6]])
  gtdn <- names(sim[[7]])
  UPTP=length(intersect(gsig_up3 ,gtup))
  UPFP=length(setdiff(gsig_up3 ,gtup))
  UPFN=length(setdiff(gtup,gsig_up3))
  DNTP=length(intersect(gsig_dn3 ,gtdn))
  DNFP=length(setdiff(gsig_dn3 ,gtdn))
  DNFN=length(setdiff(gtdn,gsig_dn3))
  TP=UPTP+DNTP
  FP=UPFP+DNFP
  FN=UPFN+DNFN
  TN=nrow(gsadn3)-DNTP-DNFP-DNFN-UPTP-UPFP-UPFN
  PREC=TP/(TP+FP)
  REC=TP/(TP+FN)
  F1=TP/(TP+(0.5*(FP+FN)))
  result <- c("TP"=TP,"FP"=FP,"FN"=FN,"TN"=TN,"PREC"=PREC,"REC"=REC)
  return(result)
}

groupsize=10
seed=100

gres <- mclapply(1:50,function(i) {
  simgsa(genesetdatabase=gs_symbols,myann=myann,gene_catalog=gene_catalog,
    mval=normal_mval,seed=i*100,frac_genes=0.5,frac_probes=0.5,groupsize=groupsize)
},mc.cores=10)
